# Confirmatory factor analysis 

In the previous section on EFA we loaded up a nice big hexaco 100 dataset and divided it in two so that we could reproduce our EFA with a CFA and not use the same observations. Let´s clean up our environment so that we can conduct our CFA on a less buisy global environment. We can use the rm() command to delete many objects like so: 

```{r message=FALSE, warning=FALSE}
rm(clean_efa_data,
   c_efa_data,
   efa_data,
   efa_cormat,
   efa.4,
   normality_diagnostics,
   resid, 
   grouping,
   factor1,
   factor2,
   factor3,
   factor4)

#Or, we can start a new R session and load in the data again like so:
library(tidyverse)
library(readxl)
library(here)
hexaco_100 <- read_excel(here("Data sets/hexaco.100.xlsx"))

set.seed(324)
grouping <- rbinom(n = 25914,size = 1, prob = .5)

cfa_data <- hexaco_100 %>%
  filter(grouping == 0)
```

For CFA we will use the lavaan package again. 

```{r message=FALSE, warning=FALSE}
library(lavaan)
library(semPlot)
```

Let´s start by selecting our data and doing some cleaning of the data using the mvn function like we did for the EFA
```{r}
#Extracting a small sample of  conscientiousness data
c_cfa_data <- cfa_data %>% dplyr::select(hex2, hex26, hex50, hex74, hex8, hex32, hex56, hex80, hex14, hex38, hex62, hex86, hex20, hex44, hex68, hex92)
set.seed(3424)
c_cfa_data <- slice_sample(c_cfa_data, n = 1000, replace = F)

library(MVN)
normality_diagnostics <- mvn(c_cfa_data
    , mvnTest = 'mardia'
    , multivariatePlot = 'qq' 
    , univariateTest = 'AD'
    , showOutliers = TRUE 
    , showNewData = TRUE 
    , multivariateOutlierMethod = 'adj')

normality_diagnostics$Descriptives
normality_diagnostics$multivariateNormality
normality_diagnostics$univariateNormality

```

Let´s take the same shortcut as we did earlier and remove the outliers from the data using the newdata from the diagnostics object. 
```{r}
 clean_cfa_data <- normality_diagnostics$newData
```

The basis for CFA is the variance covariation matrix, so lets make it! 
```{r}
covmat <- cov(clean_cfa_data)
View(covmat)
```
Looks about right, we can expect that the items within the latent factors covary more than those not encompassed by the latent variable/factor.just to confirm that we truuly have the variance covaraince matrix, lets look at the variance withing item 1(hex2), it should be the same as in the first diagonal cell of the matrix. We do this by squaring the standard deviation of the item.

```{r}
sd(clean_cfa_data$hex2)^2
```
There we have it, we are one the right track. Let´s continue by have a look at the matrix for each factor. For this we need to redefine our factors from the EFA but on the CFA data set.

```{r}
factor1 <- data.frame(clean_cfa_data$hex8, 
                      clean_cfa_data$hex32, 
                      clean_cfa_data$hex56, 
                      clean_cfa_data$hex80)

factor2 <- data.frame(clean_cfa_data$hex20, 
                      clean_cfa_data$hex44, 
                      clean_cfa_data$hex92)

factor3 <- data.frame(clean_cfa_data$hex2, 
                      clean_cfa_data$hex26, 
                      clean_cfa_data$hex50, 
                      clean_cfa_data$hex74)

factor4 <- data.frame(clean_cfa_data$hex14, 
                      clean_cfa_data$hex38, 
                      clean_cfa_data$hex62, 
                      clean_cfa_data$hex86)

cov(factor1)
cov(factor2) 
cov(factor3)
cov(factor4)
```
We see here that the vcov is is relativly high in comparison to non-related latent variable items(those approaching 0 in the large covariance matrix) - this is good. 
Now we can try to fit our model. But lets first ponder a bit on how the factors relate to eachother. Recall the factor correlation matrix from our EFA:

 	     F1  	  F2  	   F3  	   F4  
F1	 1.000	-0.436	 0.403	 0.298
F2	-0.436	 1.000	-0.386	-0.181
F3	 0.403	-0.386	 1.000	 0.215
F4	 0.298	-0.181	 0.215	 1.000

We see that some factors correlate and some do not. Let´s phrase our model accordingly. We also need to name the factors. According to the hexaco guide, our F1 is "diligence", F2 is "prudence" F3 is "organization" and F4 is "perfectionism". Thus, we see that diligence is negativly correlated with prudence and positivly correlated with organisation. We also find that prudence is negativley correlated with organisation. Let´s constrain the model so that only these covariations are allowed. That is, perfection is left uncorrelated with all other facets.

```{r}
#specifying the model

model1 <-( '

#latent variables

conscientiousness =~ diligence + prudence + organisation + perfectionism

diligence    =~ hex8 + hex32 + hex56 + hex80 
prudence     =~ hex20 + hex44 +92 
organisation =~ hex2 + hex26 + hex50 + hex74
perfectionism =~ hex14 + hex38 + hex62 + hex86

#covariates
diligence ~~ prudence
diligence ~~ organisation
prudence  ~~ organisation

')

cfa(model = model1, data = data)

m <-( '

c =~ diligence + prudence 

#latent variables
diligence =~ hex8 + hex32 + hex56 + hex80 
prudence =~ hex20 + hex44 +92')

cfa(m, clean_cfa_data)



```




